Title Goes Here

### 2020-05-09 analysis/paper/paper.Rmd
editing text

Gender under-representation and inequality has long existed, from under-representation in children’s picture books to academic publications (@hamilton2006gender; @tushingham2017peer). When talking about inequality in academic fields, we usually refer to the disproportionality between men and women in the field, including archaeology. Women are often underrepresented due to the lack of support women receive in academic fields (@xu2008gender). When focusing on the field of archaeology, we can see that gender imbalance still occurs in archaeology and that seems to impact the kind of publications that are done in the field (@bardolph2014critical). As we can see in the authorship of the main authors from the Journal of Field Archaeology, 78% of them are men (@heath2020guest). When looking at gender ratios among the membership in Society for California Archaeology from 1967 to 2016, Tushingham et al. (2017) report a trend of women increasingly maintaining their society membership and the gender gap is almost gone. In this study we investigate gender ratios of presenters at major archaeological conferences. We extend the work of Tushingham et al. (2017) with a study of gender ratios of presenters at meetings of the Society of American Archaeology (SAA, 2016-2019), the European Association of Archaeologists (EAA, 2018) and the Computer Applications and Quantitative Methods in Archaeology (CAA, 2017)  We explore how gender ratios vary over time and between these conferences. We examine the influence of the gender of the presenter on the topics that they present on.

@@

@@

@@

We collected lists of the people that presented research at recent SAA (2018 with `r SAA18_row` presenters, EAA (2018 with `r EAA18_row` presenters) and CAA (2018 with `r CAA18_row` presenters) conferences. We obtained publicly available programs of these conferences and also tidied the data into a rectangular form with variables including the primary author's first name, last name and the title of their presentation. We used R and the gender package (@lincolngender; @mihaljevic2019reflections; @blevins2015jane) to predict the gender of all first-named presenters in our sample. We used the gender package to identify a first name as male or female, based on a probability scale computed from government records of name-gender data, specifically US Social Security Administration (SSA) baby name data (@mihaljevic2019reflections). This method has the advantage of speed and ease of automation, but also some substantial limitations. We are only able to infer gender into binary male/female genders and assign probabilities for the first names into these categories. This has the unfortunate result of excluding other genders from the results, and we encourage conference organizers to collect gender data directly from participants to improve the representation of minority genders in future studies. The package often fails to classify non-English names, as the SSA data the package uses are mostly in English names. This means that people with non-English names are underrepresented in our results. To remove names where gender is ambiguous, which we define as a similar proportion of males and females having that name in the SSA data, we excluded names where the difference in proportions was less than 0.47, a figure we obtained by exploratory data analysis. It is important to note that the gender designations here are not self-identified by the presenters, but are computed probabilistically. Better quality data will come if presenters self-identify their gender, but we have no alternative because currently these data are not available. Our hope is that this work, despite its many limitations, will stimulate the collection of more reliable, justifiable, and useful data by conference organisers.

We are also interested in whether there is a variation in topic choices between these two genders in archaeological conferences. To do this we have used topic modeling, which is a method to automatically find related groups of words that resemble 'themes' or 'topics'. This approach does not treat one topic, which consists of many words, as separate from another topic. That is, a word can be present in many topics but with differs weightings. When generating topic models, we have to first decide on the number of topics that the method will identify in our texts. We then create a topic model for all our texts, this will assign all words in every document into our set number of topics, and assign each word a probability based on its per-topic-per-word probability. To visualise the topic model we use the top 10 words within each topic as keywords to represent each topic. We excluded words that are common in the field of archaeology or have little semantic value, for example, 'the', 'a', 'el', 'al', 'archaeology', etc. We generated one topic model per conference. Each presentation at each conference is then assigned a vector of topic probabilities.

We then tried another process of topic modeling called Structural Topic Model (STM) which is very similar to the LDA we have described above. The main advantage for us is that, the STM allows us to include, alongside the normal topic modeling process, other metadata. This process "attaches" the metadata to the collection of texts that we will then perform topic modeling process. It is important to note that we have used the stm package that also includes the ability to calculate the number of topics that is best suited for STM modeling (@roberts2019package). The fact that we are allowed to do this allows us to perform covariate analysis. This basically means that we can examine the relationship between the topics and some covariates/variables of interest. We can define the genders of males and females and then analyze the prevelance of these two categories in accordance with the topics that the stm package determined (@roberts2019package). The analysis is used also with probabilities given to the texts to see if the most occuring of some text seems to be "attached" to one gender over another.

To explore the relationship between gender and topics we identified the most prominent topic for each presentation...

@@

In the SAA data we found a total of `r SAA18_row` presentations. Of these we could identify the first-named author of `r count_gendered_SAA18` (`r perc_gendered_SAA18`%) presentations as either men (n=`r male_count_SAA18`) or women (n=`r female_count_SAA18`). For the CAA meeting we have `r CAA18_row` presentations, and `r count_gendered_CAA18` (`r perc_gendered_CAA18`%) first-named authors could be classified as either men (n=`r male_count_CAA18`) or women (n=`r female_count_CAA18`). For the EAA data we have `r EAA18_row` presentations, and `r count_gendered_EAA18` (`r perc_gendered_EAA18`%) where the first author could be classified as either a man (n=`r male_count_EAA18`) or woman (n=`r female_count_EAA18`). This results in similar women-to-men ratios for the SAA (`r female_to_male_SAA18`) and EAA (`r female_to_male_EAA18`), but much lower for the CAA (`r female_to_male_CAA18`).

In the SAA presentations we identified the optimum number of topics as `r SAA_stm_topic_number`, with `r SAA_stm_topic_number_significant_covary` of these showing non-random co-variance with the gender of the first-named presenter. One of the main difference that we can see between the topics of the genders are the locations that the presenters are interested in, it seems like that females are more likely to have a topic at a specific region such as mesoamerica, arizona and national parks, while males are more likely to have a topic that are in the western regions.

For EAA we have generated a total of `r EAA_stm_topic_number` topics, with `r EAA_stm_topic_number_significant_covary` of these showing non-random co-variance with the gender of the first-named presenter. Next, let's take a look at EAA's difference between females and males. Although both red-marked topics include topics with "mediterranean", it is important to distinguish that females seems to be focusing on western mediterranean while males seems to be focusing on northern mediterranean. Another main difference that we have noticed is that males tends to have topics that are related to architect, while the topic regarding research methods using isotopes seems to be more popular for females.

For CAA we hve generated a total of `r CAA_stm_topic_number`, with `r CAA_stm_topic_number_significant_covary` of these showing non-random co-variance with the gender of the first-named presenter.. Lastly, in CAA's stm topic plot we can see a clearer pattern from the topics marked in red. For females from CAA, we see that topics that are associated with Geographic Information System (GIS) are more likely to be a topic for female presenters. Another female-associated topics from the CAA topic plot is topics that includes the word women, excav and inform. As for males from CAA, there doesn't seem to have a specific word that is more associated with males besides the word enviorn.

>>>>

Gender under-representation and inequality has long existed in science, from under-representation of women and minority genders in school text books to academic publications (@hamilton2006gender; @tushingham2017peer). Women are often under-represented for several reasons, including stereotyping that unfairly underestimates individuals' abilities, which often leads to lack of support for women in academic fields (@xu2008gender). In this paper we explore the relationship between gender and research topics to determine if women and men tend to focus more on certain topics than others.

We used a computational method to identify topics from the titles of presentations delivered at meetings of the Society of American Archaeology (SAA, 2016-2019), the European Association of Archaeologists (EAA, 2018) and the Computer Applications and Quantitative Methods in Archaeology (CAA, 2017). We analysed the covariance of topics and genders of presenters to explore how gender ratios vary between these conferences and by topics.

@@

When focusing on the field of archaeology, we can see that gender imbalance still occurs in archaeology and that seems to impact the kind of publications that are done in the field (@bardolph2014critical).

As we can see in the authorship of the main authors from the Journal of Field Archaeology, 78% of them are men (@heath2020guest).

When looking at gender ratios among the membership in Society for California Archaeology from 1967 to 2016, @tushingham2017peer report a trend of women increasingly maintaining their society membership and the gender gap is almost gone. In this study we investigate gender ratios of presenters at major archaeological conferences.

@@

library(gender)
lynne_gender <-  gender("Lynne")

@@

We requested from conference organisers spreadsheets files of the publicly available program information for three major archaeology meetings held in 2018: the Society of American Archaeology (SAA, 2016-2019), the European Association of Archaeologists (EAA, 2018) and the Computer Applications and Quantitative Methods in Archaeology (CAA, 2017).

We estimated the gender of the first-named speaker for each presentation using the R programming language and the gender package (@lincolngender; @mihaljevic2019reflections; @blevins2015jane). We inferred the gender of personal names by looking each name up in the US Social Security Administration (SSA) baby name data, and calculating the overall probability that a given name was male or female (@mihaljevic2019reflections). If the proportion of people in the SSA data with a certain given name are recorded as female is 0.5 or higher, we predicted the gender of the presenter with this name as female. For example, in the SSA dataset the the name "Lynne" returns `r lynne_gender$proportion_male` as proportion of individuals with this name who are recorded as male, and `r lynne_gender$proportion_female` as the proportion female, thus we inferred that people in our conference data who are called Lynne are women.

This method of inferring gender has the advantage of speed, transparency and reproducibility, but also some substantial limitations that are important to be upfront about. We are only able to infer binary male/female genders and assign the first names into these two categories. This has the unfortunate result of excluding other genders from the results, excluding them from our analysis. We considered it impractical and invasive to write to each first-named presented to request their gender information. A further limitation of our approach is that it sometimes fails to classify non-English names at all, as the SSA data consists mostly of English names. This means that people with non-English names are underrepresented in our results. It is important to note that the inferences of gender presented here are not self-identified by the presenters, but are computed probabilistically. Better quality and more representative data would result if presenters self-identified their gender, but currently these data are not available. We encourage conference organizers to collect gender data directly from participants to improve the representation of minority genders in future studies. Our hope is that this work, despite its many limitations, will stimulate the collection of more reliable, justifiable, and useful data by conference organisers.

We identified the topic of each presentation by generating a structural topic model for all the presentations in each of the three conferences. Topic modeling is a machine learning method to automatically find related groups of words that resemble traditional 'themes' or 'topics' in a collection of documents.Topic models are 'unsupervised' methods because they infer rather than assume the content of the topics in a collection of documents, and they have been used across a variety of fields [add citations]. This approach uses  latent Dirichlet allocation to allow every word to be present in every topic, but with different weightings, such that the most heavily weighted 5-10 words of a topic often capture the essence of the topic as a coherent and familiar concept. When generating topic models, we have to first decide on the number of topics that the method will identify in our texts. The topic modeling algorithm then  assigns all words in every document into our set number of topics, and assigns each word a probability based on its per-topic-per-word probability, and each document (i.e. conference presentation title) is assigned a distribution of topics with different weights. To visualise the topic model output we use the 10 words with the highest weighting within each topic as keywords to represent each topic. In generating the topic model we excluded words that are common in English generally or in archaeology and so have little semantic value, for example, 'the', 'a', 'el', 'al', 'archaeology', etc.

A structural topic model (STM) allows us to include, alongside the typical topic modeling process, other document-level metadata to analyse covariance between topics and metadata (@roberts2019package). Covariates of interest can be included into the prior distributions for document-topic proportions and
topic-word distributions using a standard regression model (cite this one: http://scholar.harvard.edu/files/dtingley/files/topicmodelsopenendedexperiments.pdf). This means that we can examine the relationship between the topics and document-level variables of interest. In our case, we computed the relationship between the inferred the gender of the first-named presenter for each presentation, and the distribution of topics in each presentation. When we observed the topics for each presentation, we generated a regression where the topic is the outcome variable, and gender is the explanatory variable. This regression gives us insights into whether gender caused archaeologists to spend a larger portion of their presentation on a particular topic.

@@

In the SAA data we found a total of `r prettyNum(SAA18_row, big.mark = ",")` presentations. Of these we could identify the first-named author of `r count_gendered_SAA18` (`r perc_gendered_SAA18`%) presentations as either men (n=`r male_count_SAA18`) or women (n=`r female_count_SAA18`). For the CAA meeting we have `r prettyNum(CAA18_row, big.mark = ",")` presentations, and `r count_gendered_CAA18` (`r perc_gendered_CAA18`%) first-named authors could be classified as either men (n=`r male_count_CAA18`) or women (n=`r female_count_CAA18`). For the EAA data we have `r prettyNum(EAA18_row, big.mark = ","` presentations, and `r count_gendered_EAA18` (`r perc_gendered_EAA18`%) where the first author could be classified as either a man (n=`r male_count_EAA18`) or woman (n=`r female_count_EAA18`). This results in similar women-to-men ratios for the SAA (`r female_to_male_SAA18`) and EAA (`r female_to_male_EAA18`), but much lower for the CAA (`r female_to_male_CAA18`).

In the SAA presentations we identified the optimum number of topics as `r SAA_stm_topic_number`, with `r SAA_stm_topic_number_significant_covary` of these showing non-random co-variance with the gender of the first-named presenter. One of the main difference that we can see between the topics of the genders are the locations that the presenters are interested in. Women are more likely to have a topic at a specific region such as Mesoamerica, Arizona and national parks, while men are more likely to have a topic that are in the western regions.

For EAA we generated a total of `r EAA_stm_topic_number` topics, with `r EAA_stm_topic_number_significant_covary` of these showing non-random co-variance with the gender of the first-named presenter. Although topics  significantly associated with men and women both include the word "Mediterranean", it is important to distinguish that females seems to be focusing on western Mediterranean while males seems to be focusing on northern Mediterranean Another  difference  is that males tends to have topics that are related to architecture, while the topic regarding research methods using isotopes seems to be more popular for females.

For CAA we generated a total of `r CAA_stm_topic_number` topics, with `r CAA_stm_topic_number_significant_covary` of these showing non-random co-variance with the gender of the first-named presenter. In CAA's stm topic plot we can see a clearer pattern from the topics marked in red. For females from CAA, we see that topics that are associated with Geographic Information System (GIS) are more likely to be a topic for female presenters. Another female-associated topics from the CAA topic plot is topics that includes the word women, excav and inform. As for males from CAA, there doesn't seem to have a specific word that is more associated with males besides the word enviorn.
